# Implementation of LGCP with INLA and SPDE

To install INLA we run
```{r}
#| eval: false
options(repos = c(getOption("repos"), INLA = "https://inla.r-inla-download.org/R/testing"))
# Install INLA and dependencies
install.packages("INLA", dependencies = TRUE)
remotes::install_version("INLA", version="23.04.24",repos=c(getOption("repos"),INLA="https://inla.r-inla-download.org/R/testing"), dep=TRUE)

if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("Rgraphviz")
# This one
remotes::install_version("INLA", version="24.05.01-1",repos=c(getOption("repos"),INLA="https://inla.r-inla-download.org/R/stable"), type="binary", dep=TRUE)

Sys.getenv("PATH")

# In the end I downloaded the INLA zip INLA_24.05.01-1.zip and did from https://inla.r-inla-download.org/R/stable/bin/windows/contrib/4.3/
# I was getting a cp error and I don't know why, if it was related to Rtools path or what. 
install.packages("C:/Users/msavery/Downloads/INLA_24.05.01-1.zip", repos = NULL, type = "win.binary", lib="C:/Program Files/R/R-4.3.3/library")

```

```{r}
#| eval: false
library(viridis)
library(hrbrthemes)
library(ggplot2)
#library(cmdstanr) 
#library(bayesplot)
library(tidyterra)
library(terra)
library(INLA)
library(inlabru)
#source("stan_models/lgcp.R")
# TODO: Add these files
#source("presence_only_functions.R")
#source("surface_functions.R")
```
<https://www.youtube.com/watch?v=a8QvxCjWieg> lecture here Havard Rue says "If you go through the math, it's too good to be true. Everything you hope for turns out to be true." 

@lindgren2015 is the software paper for INLA.

To install the R-INLA package, run "install.packages("INLA",repos=c(getOption("repos"),INLA="https://inla.r-inla-download.org/R/stable"), dep=TRUE)"

Here we are going to cover the basics of the log-Gaussian Cox Process (INLA), it's approximation using the Integrated Nested Laplace Approximation, and the application of this model to Presence-Only point pattern modelling.

There are many good references, as this is a somewhat trendy model in ecology.
- https://www.r-inla.org/examples-tutorials   
- https://akawiecki.github.io/statistical_rethinking_inla/index.html
- https://www.precision-analytics.ca/articles/a-gentle-inla-tutorial/#1-bayesian-inference    
- https://www.pymc.io/projects/docs/en/v3.11.4/pymc-examples/examples/case_studies/log-gaussian-cox-process.html   

Below is taken from <https://becarioprecario.bitbucket.io/spde-gitbook/ch-INLA.html>.
```{r}
#| label: inla-fit
#fit the model
data(SPDEtoy)
SPDEtoy.sp <- SPDEtoy
coordinates(SPDEtoy.sp) <- ~ s1 + s2

bubble(SPDEtoy.sp, "y", key.entries = c(5, 7.5, 10, 12.5, 15), 
  maxsize = 2, xlab = "s1", ylab = "s2")
#dat.inla <- inla(y~x+f(unit, model='iid'), family='poisson',
#   data=dat.pred,
#   control.family=list(link='log'),
#   control.predictor=list(link=1, compute=TRUE),
#   control.compute=list(dic=TRUE, cpo=TRUE, waic=TRUE))
m0 <- inla(y ~ s1 + s2, data = SPDEtoy)
#examine the regular summary 
summary(m0)
```
Then look at an inlabru example, from <https://inlabru-org.github.io/inlabru/articles/2d_lgcp_covars.html>

How do we use inlabru? Depends on if we are using SP or SF data structures
```{r inlabru}
data(gorillas_sf, package = "inlabru")
nests <- gorillas_sf$nests
mesh <- gorillas_sf$mesh
boundary <- gorillas_sf$boundary
gcov <- gorillas_sf_gcov()
ggplot() +
  gg(gcov$vegetation) +
  gg(boundary, alpha = 0.2) +
  gg(nests, color = "white", cex = 0.5)
# with the mesh
ggplot() +
  gg(gcov$vegetation) +
  gg(mesh) +
  gg(boundary, alpha = 0.2) +
  gg(nests, color = "white", cex = 0.5)
comp1 <- geometry ~ vegetation(gcov$vegetation, model = "factor_full") - 1
comp1alt <- geometry ~ vegetation(gcov$vegetation, model = "factor_contrast") + Intercept(1)
fit1 <- lgcp(comp1, nests, samplers = boundary, domain = list(geometry = mesh))
pred.df <- fm_pixels(mesh, mask = boundary)
int1 <- predict(fit1, pred.df, ~ exp(vegetation))

ggplot() +
  gg(int1, geom = "tile") + # gg() with sf points and geom = "tile" plots a raster
  gg(boundary, alpha = 0, lwd = 2) +
  gg(nests, color = "DarkGreen")
```
And show the example with sp data structure.
```{r sp-inlabru}
data(gorillas, package = "inlabru")
nests <- gorillas$nests
mesh <- gorillas$mesh
boundary <- gorillas$boundary
ggplot() +
  geom_fm(data = mesh) +
  gg(nests) +
  gg(boundary, fill = "red", alpha = 0.2) +
  ggtitle("Points")
matern <- inla.spde2.pcmatern(
  mesh,
  prior.sigma = c(0.1, 0.01),
  prior.range = c(5, 0.01)
)

nests
class(nests)
cmp <- coordinates ~
  mySmooth(coordinates, model = matern) +
  Intercept(1)

fit <- lgcp(cmp, nests, samplers = boundary, domain = list(coordinates = mesh))
```
Look at a spacetime example, from <https://inlabru-org.github.io/inlabru/articles/2d_lgcp_spatiotemporal.html>.
The temporal aspect is a bit simple: Just some sort of indicator for one of four seasons.
```{r space-time}
data(mrsea, package = "inlabru")
ggplot() +
  gg(mrsea$mesh) +
  gg(mrsea$boundary) +
  gg(mrsea$samplers) +
  gg(mrsea$points, size = 0.5) +
  facet_wrap(~season) +
  ggtitle("MRSea observation seasons")
# Separate integration meshes for each season

ips <- fm_int(
  domain = list(coordinates = mrsea$mesh, season = 1:4),
  samplers = mrsea$samplers
)
ggplot() +
  gg(mrsea$mesh) +
  gg(ips, aes(size = weight)) +
  scale_size_area(max_size = 1) +
  facet_wrap(~season)
matern <- inla.spde2.pcmatern(mrsea$mesh,
  prior.sigma = c(0.1, 0.01),
  prior.range = c(10, 0.01)
)

cmp <- coordinates + season ~ Intercept(1) +
  mySmooth(
    coordinates,
    model = matern,
    group = season,
    ngroup = 4
  )

fit <- lgcp(cmp,
  data = mrsea$points,
  samplers = mrsea$samplers,
  domain = list(
    coordinates = mrsea$mesh,
    season = seq_len(4)
  )
)
summary(fit)
ppxl <- fm_pixels(mrsea$mesh, mask = mrsea$boundary, format = "sp")
ppxl
ggplot() +
geom_sf(data=as(ppxl, "sf"))

ppxl_all <- fm_cprod(ppxl, data.frame(season = seq_len(4)))
ppxl_all
plot(ppxl_all)
class(ppxl_all)
lambda1 <- predict(
  fit,
  ppxl_all,
  ~ data.frame(season = season, lambda = exp(mySmooth + Intercept))
)
lambda1
pl1 <- ggplot() +
  gg(as(lambda1, "SpatialPixelsDataFrame"), aes(fill = q0.5)) +
  gg(mrsea$points, size = 0.3) +
  facet_wrap(~season) +
  coord_equal()
pl1
```
Finally look at the example from <https://becarioprecario.bitbucket.io/spde-gitbook/ch-spacetime.html#discrete-time-domain>.

Another interesting example is here <https://becarioprecario.bitbucket.io/inla-gitbook/ch-temporal.html#sec:spacetime>. In the section [here]<https://becarioprecario.bitbucket.io/inla-gitbook/ch-temporal.html#tab:controlgroup> there is a good example of how to provide arguments for the group option, the temporal index.

I would like to have this done in INLABRU. More specifically, the model of Cameletti 2013, with autoregressive time + spatial covariance conditional on time $t$.

As of 12-08-2024, code is not working due to some deprecations.
```{r discrete-time-space}

# Book function copied from spde-book-functions.R
load('data/prmesh1.RData')
book.rspde <- function(coords, sigma=1, range, variance=sigma^2, alpha=2, kappa = sqrt(8*(alpha-1))/range, n=1, mesh, 
                  verbose=FALSE, seed, return.attributes=FALSE) {
    t0 <- Sys.time()
    theta <- c(-0.5*log(4*pi*variance*kappa^2), log(kappa))
    if (verbose) cat('theta =', theta, '\n')
    if (missing(mesh)) {
        mesh.pars <- c(0.5, 1, 0.1, 0.5, 1)*sqrt(alpha-ncol(coords)/2)/kappa 
        if (verbose) cat('mesh.pars =', mesh.pars, '\n')
        attributes <- list(
            mesh=inla.mesh.2d(,
                coords[chull(coords), ], max.edge=mesh.pars[1:2], 
                cutoff=mesh.pars[3], offset=mesh.pars[4:5]))
        if (verbose) cat('n.mesh =', attributes$mesh$n, '\n')
    }
    else attributes <- list(mesh=mesh)
    attributes$spde <- inla.spde2.matern(attributes$mesh, alpha=alpha)
    attributes$Q <- inla.spde2.precision(attributes$spde, theta=theta)
    attributes$A <- inla.mesh.project(mesh=attributes$mesh, loc=coords)$A
    if (n==1) 
        result <- drop(attributes$A%*%inla.qsample(
            Q=attributes$Q,
            constr=attributes$spde$f$extraconstr))
    t1 <- Sys.time() 
    result <- inla.qsample(n, attributes$Q, 
                           seed=ifelse(missing(seed), 0, seed), 
                           constr=attributes$spde$f$extraconstr) 
    if (nrow(result)<nrow(attributes$A)) {
        result <- rbind(result, matrix(
            NA, nrow(attributes$A)-nrow(result), ncol(result)))
        dimnames(result)[[1]] <- paste('x', 1:nrow(result), sep='')
        for (j in 1:ncol(result)) 
            result[, j] <- drop(attributes$A%*%
                                result[1:ncol(attributes$A),j])
    }
    else {
        for (j in 1:ncol(result)) 
            result[1:nrow(attributes$A), j] <-
                drop(attributes$A%*%result[,j]) 
        result <- result[1:nrow(attributes$A), ]
    }
    t2 <- Sys.time()
    attributes$cpu <- c(prep=t1-t0, sample=t2-t1, total=t2-t0)
    if (return.attributes) 
        attributes(result) <- c(attributes(result), attributes)
    return(drop(result))
}

data(PRborder)
k <- 12
coords <- as.matrix(PRprec[sample(1:nrow(PRprec)), 1:2])
params <- c(variance = 1, kappa = 1) 

set.seed(1)
x.k <- book.rspde(coords, range = sqrt(8) / params[2], 
  sigma = sqrt(params[1]), n = k, mesh = prmesh1,
  return.attributes = TRUE)
rho <- 0.7
x <- x.k
for (j in 2:k) 
  x[, j] <- rho * x[, j - 1] + sqrt(1 - rho^2) * x.k[, j]
n <- nrow(coords)
set.seed(2)
ccov <- factor(sample(LETTERS[1:3], n * k, replace = TRUE))
beta <- -1:1
sd.y <- 0.1
y <- beta[unclass(ccov)] + x + rnorm(n * k, 0, sd.y)
isel <- sample(1:(n * k), n * k / 2) 
dat <- data.frame(y = as.vector(y), w = ccov, 
  time = rep(1:k, each = n), 
  xcoo = rep(coords[, 1], k), 
  ycoo = rep(coords[, 2], k))[isel, ] 
spde <- inla.spde2.pcmatern(mesh = prmesh1, 
  prior.range = c(0.5, 0.01), # P(range < 0.05) = 0.01
  prior.sigma = c(1, 0.01)) # P(sigma > 1) = 0.01
iset <- inla.spde.make.index('i', n.spde = spde$n.spde,
  n.group = k)
A <- inla.spde.make.A(mesh = prmesh1,
  loc = cbind(dat$xcoo, dat$ycoo), group = dat$time) 
sdat <- inla.stack(
  data = list(y = dat$y), 
  A = list(A, 1), 
  effects = list(iset, w = dat$w),
  tag = 'stdata') 
h.spec <- list(rho = list(prior = 'pc.cor1', param = c(0, 0.9)))
# Model formula
formulae <- y ~ 0 + w + f(i, model = spde, group = i.group, 
  control.group = list(model = 'ar1', hyper = h.spec)) 
# PC prior on the autoreg. param.
prec.prior <- list(prior = 'pc.prec', param = c(1, 0.01))
# Model fitting
res <- inla(formulae,  data = inla.stack.data(sdat), 
  control.predictor = list(compute = TRUE,
    A = inla.stack.A(sdat)), 
  control.family = list(hyper = list(prec = prec.prior)), 
  control.fixed = list(expand.factor.strategy = 'inla'))
cbind(Obs. = tapply(dat$y, dat$w, mean), res$summary.fixed[, -7])
idat <- inla.stack.index(sdat, 'stdata')$data
cor(dat$y, res$summary.linear.predictor$mean[idat])
stepsize <- 4 * 1 / 111
nxy <- round(c(diff(range(coords[, 1])), 
  diff(range(coords[, 2]))) / stepsize)
projgrid <- inla.mesh.projector(
  prmesh1, xlim = range(coords[, 1]), 
  ylim = range(coords[, 2]), dims = nxy)
xmean <- list()
for (j in 1:k)
  xmean[[j]] <- inla.mesh.project(
    projgrid, res$summary.random$i$mean[iset$i.group == j])
library(splancs)
xy.in <- inout(projgrid$lattice$loc, 
  cbind(PRborder[, 1], PRborder[, 2]))
```
